\section*{Computational Methods}

The (111) metal surfaces were modeled with slabs containing 192 metal atoms arranged in four 8 $\times$ 6 atomic layers. \SI{10}{\angstrom} of vacuum was added in the direction normal to the surface to ensure weak interaction between periodic images of the slab. The size of the slabs in the lateral directions was $20.55 \times 13.35$~\si{\angstrom} for Cu, $22.84 \times 14.82$~\si{\angstrom} for Ag, and $22.47 \times 14.60$~\si{\angstrom} for Au. When pre-existing atoms are considered, it is assumed that the presence of the vacancy does not affect the state energetics drastically and, therefore, the vacancy-containing models are re-used to model pre-existing adatoms. 

DFT calculations were performed using Vienna \emph{ab initio} simulation package (VASP)~\cite{ullmann_131, ullmann_132, ullmann_133, ullmann_134}. The dispersion-corrected~\cite{ullmann_136, ullmann_137} Perdew-Burke-Ernzerhof (PBE) generalized gradient approximation~\cite{ullmann_139} was used as the exchange-correlation functional. 
Spin-polarized electronic states were modeled using a plane wave basis set with the energy cut-off set at \SI{800}{\electronvolt}.
The projector augmented wave formalism was used to describe interactions of atomic cores with valence electrons. The integration over the Brillouin zone was performed using the $3\times 3 \times1$ Monkhorst-Pack $k$-point mesh. 

Atomic positions were optimized until the maximum force on atoms decreased below \SI{0.02}{\electronvolt\per\angstrom}. 
Transition state structures were located using the climbing-image nudged elastic band (NEB) with the VTST code~\cite{ullmann_59}. 
In NEB calculations, an improved initial guess~\cite{ullmann_60, ullmann_99} for the minimum energy path was used and the positions of atoms were relaxed until the maximum force dropped below \SI{0.1}{\electronvolt\per\angstrom}.

{\zhzh In addition, the effect of entropy was also verified in the process of Ullmann reaction. Compared electronic energy from DFT calculation and Gibbs free energy after applying entropy correction, the effect is estimated to play a minor role in Ullmann process, details of results in ~\ref{SI-fig:entropy_ideal} and \ref{SI-fig:entropy_adatom}}.

%{\comm RZK0804: Energy vs enthalpy.}
